\section*{Acknowledgments}

We thank the Cornell Cayuga HPC team for computational resources and technical
support. We acknowledge the Arc Institute and the Evo2 development team for
making the model checkpoint publicly available. We thank the ClinVar, MaveDB,
ENCODE, and gnomAD teams for generating and maintaining the datasets used in this
study.

\subsection*{Funding}

This work was supported by the NASA GeneLab project and the WorldQuant Foundation
through grants to C.E.M.

\subsection*{Author Contributions}

J.K.\ conceived the analytical framework, developed all analysis pipelines, ran
all computations, and drafted the manuscript. C.E.M.\ supervised the project,
provided scientific direction, and critically revised the manuscript. Both
authors approved the final version.

\subsection*{Data Availability}

Pre-computed Evo2 $\Delta$LL scores for all 215,001 variants, ClinVar benchmarking
results, DMS calibration data, and all analysis outputs are deposited in the
GitHub repository at \url{https://github.com/JangKeunKim/evo2-spaceflight-vep}.

ClinVar variant data are available from NCBI
(\url{https://www.ncbi.nlm.nih.gov/clinvar/}).
DMS and MPRA datasets are available from MaveDB~\cite{MaveDB_2019}
(\url{https://mavedb.org}).
ENCODE cCRE annotations are available from the ENCODE Data Coordination Center
(\url{https://screen.encodeproject.org}).
AlphaMissense, CADD v1.7, REVEL, SpliceAI, and ncER scores were obtained from
their respective public repositories.

\subsection*{Code Availability}

All analysis code, including variant generation, Evo2 scoring, benchmarking, and
figure generation scripts, is available at
\url{https://github.com/JangKeunKim/evo2-spaceflight-vep}
under the MIT License.

\subsection*{Competing Interests}
The authors declare no competing interests.
